The cost of school consolidations: impacts of reduced accessibility

### 2024-04-10 paper/Manuscript.qmd
Initialized renv and reorganized the repo: the paper, bibliography, and figures now have their own folders

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
format: pdf
editor: source
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University
    footnote: "Corresponding Author"
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University
address:
  - code: McMaster University
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
    | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes. Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

<!--- 2016 and 2011 sp_avail is calculated in 'test-doc-2016run', LOAD THE BELOW! -->
```{r, include=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
library(scales)
library(stplanr)
library(tmap)
```

```{r, include=FALSE}
load(file="data-prep/data-products/cen_data_2011_hamiltononly.Rdata")
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(file="data-prep/data-products/cen_data_2016_hamiltononly.Rdata")
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")

ham_city_bound <- st_read("data-prep/data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read("data-prep/data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

@@

```{r, include=FALSE}
library(hamONdests)
data(Schools_2011_2016)
ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
rm("Schools_2011_2016")
data(School_Catchments_2011_2016)
ALL_SCHOOLS_CATCH_Elem <- School_Catchments_2011_2016 %>% filter(Level == "Elementary")
rm("School_Catchments_2011_2016")

load(file="data-prep/data-products/TTS_gamma2011_motor.Rdata")
load(file="data-prep/data-products/TTS_gamma2016_motor.Rdata")
load(file="data-prep/data-products/TTS_exp2011_nonmotor.Rdata")
load(file="data-prep/data-products/TTS_exp2016_nonmotor.Rdata")
```

@@

tmap_save(catchment_schools_plot, "Figures/catchment_schools_plot.png", height=7, width=12, dpi=600)
```

```{r,include=FALSE}
load(file="data-prep/data-products/avail_2011.RData")
load(file="data-prep/data-products/avail_2016.RData")
```

@@

tmap_save(stud_pop_rate_LIM_plot, "Figures/stud_pop_rate_LIM_plot.png", height=7, width=12, dpi=600)
```

```{r importing-taz-and-associated-flows}
#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_taz <- st_read("data-prep/data-inputs/TTS/tts06_83_region.shp") %>% st_transform(crs = 32617)

@@

#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_pd <- st_read("data-prep/data-inputs/TTS/tts06_pd_83.shp") %>%
  st_transform(crs = 32617)

@@

```{r include=FALSE}
#the tts origin-destination trips, by mode
load(file="data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata")
load(file="data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata")

@@

walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
      tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = c(0.8,0,0)) +
  tm_add_legend(type = "line", 
                col = c("grey40","lightblue"),
                labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")
OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
tmap_save(OD_flow_plot, "Figures/OD_flow_plot.png", height=7, width=12, dpi=600)
```

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() + 
  geom_histogram(data = od_trips_perc_tt,
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey80",
                 alpha = 0.5,
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8) + 
  xlab("Estimated TTS travel time (minutes)") +
  scale_x_continuous(breaks = seq(0,40,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  theme_classic()
TLD_empirical_wtheo_curve_plot
ggsave("Figures/TLD_empirical_wtheo_curve_plot.png", height=7, width=12, dpi=600)
```

@@

tmap_save(Vim_vvim_20112016_plot, "Figures/Vim_vvim_20112016_plot.png", height=7, width=12, dpi=600)
```

@@

tmap_save(Vim_vvim_20112016_CHANGE_plot, "Figures/Vim_vvim_20112016_CHANGE_plot.png", height=7, width=12, dpi=600)
```

```{r, eval=FALSE}
walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
      tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = c(0.8,0,0)) +
  tm_add_legend(type = "line", 
                col = c("grey40","lightblue"),
                labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")
OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
tmap_save(OD_flow_plot, "Figures/SA-and-SAperStud-20112016.png", height=7, width=12, dpi=600)
```

@@

knitr::include_graphics("Figures/catchment_schools_plot.png")
```

@@

```{r elem-res-rate-2016, fig.cap="\\label{fig:Fig2}The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016. The proportion of low-income households with children under 18 years old is also shown (bottom row). All scales are represented in quartiles.", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/stud_pop_rate_LIM_plot.png")
```

@@

```{r imped-20112016, fig.cap="\\label{fig:Fig3}The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries. ", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/OD_flow_plot.png")
```
Furthermore, as a sixth data source, an estimated travel time matrix for walk and car travel modes is calculated using {r5r} and the geographical centroids of the TTS origin-destination flows.  For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022]. The walk travel times are filtered to only include travel times under 27 minutes. These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics (included in Appendix), the gamma and exponential distribution is selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.

```{r imped-20112016-2, fig.cap="\\label{fig:Fig4}The empirical and theoretical motorized and non-motorized impedance functions. ", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/TLD_empirical_wtheo_curve_plot.png")
```

@@

```{r Vim-vim-4-plots, fig.cap="\\label{fig:Fig6}Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016.", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/Vim_vvim_20112016_plot.png")
```

@@

```{r change-Vim-vim-plot, fig.cap="\\label{fig:Fig7}The change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census.", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/Vim_vvim_20112016_CHANGE_plot.png")
```

>>>>

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format: pdf
editor: source
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University
    footnote: "Corresponding Author"
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University
address:
  - code: McMaster University
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
    | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes. Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

<!--- 2016 and 2011 sp_avail is calculated in 'test-doc-2016run', LOAD THE BELOW! -->
```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
```

```{r, include=FALSE}
load(file = paste0(here::here(),"/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(file=paste0(here::here(),"/data-prep/data-products/cen_data_2016_hamiltononly.Rdata"))
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")

ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

@@

```{r, include=FALSE}
library(hamONdests) # A R package of Hamilton, Ontario destinations
data(Schools_2011_2016)
ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
rm("Schools_2011_2016")
data(School_Catchments_2011_2016)
ALL_SCHOOLS_CATCH_Elem <- School_Catchments_2011_2016 %>% filter(Level == "Elementary")
rm("School_Catchments_2011_2016")

load(file = paste0(here::here(),"/data-prep/data-products/TTS_gamma2011_motor.Rdata"))
load(file = paste0(here::here(),"/data-prep/data-products/TTS_gamma2016_motor.Rdata"))
load(file = paste0(here::here(),"/data-prep/data-products/TTS_exp2011_nonmotor.Rdata"))
load(file = paste0(here::here(),"/data-prep/data-products/TTS_exp2016_nonmotor.Rdata"))
```

@@

tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/catchment_schools_plot.png"), height=7, width=12, dpi=600)
```

```{r,include=FALSE}
load(file = paste0(here::here(),"/data-prep/data-products/avail_2011.RData"))
load(file = paste0(here::here(),"/data-prep/data-products/avail_2016.RData"))
```

@@

tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=7, width=12, dpi=600)
```

```{r importing-taz-and-associated-flows}
#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp")) %>% st_transform(crs = 32617)

@@

#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp")) %>%
  st_transform(crs = 32617)

@@

```{r include=FALSE}
#the tts origin-destination trips, by mode
load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata"))
load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata"))

@@

walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
      tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = c(0.8,0,0)) +
  tm_add_legend(type = "line", 
                col = c("grey40","lightblue"),
                labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")
OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/OD_flow_plot.png"), height=7, width=12, dpi=600)
```

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() + 
  geom_histogram(data = od_trips_perc_tt,
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey80",
                 alpha = 0.5,
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8) + 
  xlab("Estimated TTS travel time (minutes)") +
  scale_x_continuous(breaks = seq(0,40,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  theme_classic()
TLD_empirical_wtheo_curve_plot
ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=12, dpi=600)
```

@@

tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=7, width=12, dpi=600)
```

@@

tmap_save(Vim_vvim_20112016_CHANGE_plot, 
          paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"),
          height=7,
          width=12, 
          dpi=600)
```

```{r, eval=FALSE}
walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
      tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = c(0.8,0,0)) +
  tm_add_legend(type = "line", 
                col = c("grey40","lightblue"),
                labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")
OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
tmap_save(OD_flow_plot, 
          paste0(here::here(),"/Figures/SA-and-SAperStud-20112016.png"),
          height=7, 
          width=12,
          dpi=600)
```

@@

knitr::include_graphics(paste0(here::here(),"/Figures/catchment_schools_plot.png"))
```

@@

```{r elem-res-rate-2016, fig.cap="\\label{fig:Fig2}The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016. The proportion of low-income households with children under 18 years old is also shown (bottom row). All scales are represented in quartiles.", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"))
```

@@

```{r imped-20112016, fig.cap="\\label{fig:Fig3}The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries. ", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/OD_flow_plot.png"))
```
Furthermore, as a sixth data source, an estimated travel time matrix for walk and car travel modes is calculated using {r5r} and the geographical centroids of the TTS origin-destination flows.  For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022]. The walk travel times are filtered to only include travel times under 27 minutes. These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics (included in Appendix), the gamma and exponential distribution is selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.

```{r imped-20112016-2, fig.cap="\\label{fig:Fig4}The empirical and theoretical motorized and non-motorized impedance functions. ", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
```

@@

```{r Vim-vim-4-plots, fig.cap="\\label{fig:Fig6}Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016.", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
```

@@

```{r change-Vim-vim-plot, fig.cap="\\label{fig:Fig7}The change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census.", out.width="100%", echo=FALSE}
knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
```
